---
title: "Basics of R"
author: "Dr. Searcy"
date: "2023-06-22"
output: html_document
editor_options: 
  chunk_output_type: console
---

Welcome to RStudio. This is your typical interface for the program. The top left panel (where you currently are) is the R markdown file where you can add notes and code simultaneously. The top right panel has 5 different tabs: Environment, History, Connections, Git, and Tutorial. The Environment will list any saved variables that you have during your R session; this will be the most useful for you. The bottom left panel has the console. This is where your code will run when you try to execute it. The bottom right panel provides a lot of information about your current R session. There are 6 tabs in this panel: Files, Plots, Packages, Help, Viewer, and Presentation. The files has a list of files found within your current working directory. The plots tab will show the figures that you produce via code. The packages tab shows a list of all packages installed and/or activated in your R session. The help page shows information on any functions that you search for. You can ignore the viewer and presentation tabs. 

You are using what we call an R Markdown file. This file type allows you to add notes in addition to executable code. Any code that you want to run needs to be in what we call a code chunk. Below is an example of a code chunk. 

```{r}

```

If there is anything inside of the chunk that you do not want to run, then you put a # sign in front of it. 
```{r}
# Like This <- this line would not run
```

Inside of a chunk, you can execute code. To execute an entire chunk, click on the green triangle in the top right of the trunk. To runn a single line in the code chunk, highlight the line then press CMD Enter at the same time on the keyboard. R follows basic mathematical equations like so: 
```{r}
# Addition
2 + 2

# Subtraction
5 - 3

# Multiplication
84 * 16

# Division
44 / 11

# Exponents
2 ** 2
```

Sometimes you need to save a variable so that you can reference it later. You can set the variable using the `=` or the `<-` symbols. When you run this, you will see the variable show up in the Environment tab in the top right panel. 
```{r}
x <- 2

# OR

x = 2
```

Since you have saved a variable, you can recall the saved value by typing the variable name. If you run this line, you will see a value of 2 show up in the console below.  
```{r}
x
```
You can complete calculations with these variables: 
```{r}
x + 53

# Multiple Variables: 

y = 568

x + y
```

Data types: 
```{r}
# Character data: 
letters <- c("A", "B", "C", "D", "E")

# Numeric data: 
numbers <- c(1, 2, 3, 4, 5)

# Integer data: 
integers <- as.integer(numbers)

# Factor data: 
factors <- as.factor(letters)
```

Calculations with data lists:
```{r}
# Addition of data lists
numbers + integers

# Subtraction of data lists
numbers - integers

# Multiplication of data lists
numbers * integers

# Division of data lists
numbers / integers

# Exponents of data lists
numbers ** integers
```

Loading outside data: 

Reading .csv files
```{r}
Torn.dat <- read.csv("1950-2022_actual_tornadoes.csv")
```

Reading .RData files
```{r}
load("TornEnv.RData")
```

Accessing components of the data frame: 
```{r}
# Get the dimensions of the data frame

dim(Torn.dat)

# 68701 rows
# 29 columnns
```

```{r}
#get list of column names and data types: 

str(Torn.dat)
```

```{r}
# Get the top 10 rows of the data frame: 

head(Torn.dat)

# Get the bottom 10 rows of the data frame: 

tail(Torn.dat)
```
```{r}
# Access a column in the data frame using a name of the column: 

Torn.dat$mag #magnitude of the tornadoes

#Access a row

Torn.dat[1, ] # Row 1 in the data frame

# Access a specific value in a known row and column

Torn.dat$mag[50] # OR
Torn.dat[50, 11] # mag is the 11 column
```
## Installing and Loading Packages: 

**It MUST have quotation marks around the package name.**
```{r}
install.packages("dplyr")
```

Accessing the functions in an installed package:
```{r}
library(dplyr)

# OR

library("dplyr")
```

## Manipulating data in a data frame:

Functions in dplyr package: 

Pipe symbol %>%

`summarize` function is a great way to make tables of information about your data!
```{r}
Torn.dat %>%
  group_by(mag) %>% #groups data by the magnitude column
  summarize(totalcounts = n()) #calculates the number of tornadoes in each group (i.e., mag)
```

```{r}
# Filter tornadoes with a magnitude greater than EF 4
Torn.dat %>%
  filter(mag > 4)
# Filter tornadoes with a magnitude EF 4 and greater
Torn.dat %>%
  filter(mag >= 4)
# Filter tornadoes with a magnitude less than EF 4
Torn.dat %>%
  filter(mag < 4)
# Filter tornadoes with a magnitude EF 4 and less
Torn.dat %>%
  filter(mag <= 4)
# Filter tornadoes with a magnitude of EF4
Torn.dat %>%
  filter(mag == 4)
# Filter tornadoes with a magnitude not equal to EF4
Torn.dat %>%
  filter(mag != 4)
```

```{r}
Torn.dat %>%
  arrange(mag) # least to greatest

Torn.dat %>%
  arrange(desc(mag)) # greatest to least
```
`mutate` creates a new column in the data frame. In order to save the added columns in the data frame, you must re-save the object. 
```{r}
Torn.dat <- Torn.dat %>%
  mutate(cas = inj + fat)
```

## Making figures with ggplot2: 

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

Get USA data using a specialty package from github: 
```{r}
#remotes::install_github("ropensci/USAboundaries")
#remotes::install_github("ropensci/USAboundariesData")
library(USAboundaries)
```

Names of datasets: 
* us_cities
* us_congressional
* us_counties
* us_states
* us_zipcodes
```{r}
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")] #Removes AK and HI from list of states
stateBorders <- us_states(states = sts)
```

**Most Used**
geom_line         geom_histogram
geom_label        geom_text
geom_bar          geom_point
geom_ribbon       geom_abline
geom_hline        geom_boxplot
geom_polygon
```{r}
#Set up data for a bar graph
dat <- Torn.dat %>%
  group_by(yr) %>%
  summarize(nT = n())
```

As a bar graph
```{r}
ggplot(dat, aes(x = yr, y = nT)) + 
  # Basic bargraph with year on the x axis and number of tornadoes on the y axis
  geom_bar(stat = "identity", 
           position = "stack", 
           fill = "black",
           width = 0.9) +
  # Add a regression trend line
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              se = FALSE,
              color = "red",
              linewidth = 1.5) +
  # Add a simple black and white background for the bar graph
  theme_bw() +
  # Change the axis labels
  xlab("Decade") +
  ylab("Number of Tornadoes") +
  # Change the text size on the axis ticks and the axis labels
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16), 
        legend.text = element_text(size = 16)) +
  # Adjust the limits, number, and spacing of the tickmarks on the y axis 
  scale_y_continuous(expand = c(0, 0),  
                     limits=c(0, 1850), 
                     breaks = seq(0, 1850, 250)) +
  # Adjust the limits, number, and spacing of the tickmarks on the x axis 
  scale_x_continuous(expand = c(0, 0), 
                     limits=c(1950, 2023), 
                     breaks = seq(1950, 2023, 10))
```

As a line graph:
```{r}
ggplot(dat, aes(x = yr, y = nT)) + 
  # Basic line graph with year on the x axis and number of tornadoes on the y axis
  geom_line(stat = "identity", 
           position = "stack") +
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              se = FALSE,
              color = "red") +
  theme_bw() +
  xlab("Decade") +
  ylab("Number of Tornadoes") +
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16), 
        legend.text = element_text(size = 16)) +
  scale_y_continuous(expand = c(0, 0),  
                     limits=c(0, 1850), 
                     breaks = seq(0, 1850, 250)) +
  scale_x_continuous(expand = c(0, 0), 
                     limits=c(1950, 2023), 
                     breaks = seq(1950, 2023, 10))
```

As a dot plot
```{r}
ggplot(dat, aes(x = yr, y = nT)) + 
  # Basic bargraph with year on the x axis and number of tornadoes on the y axis
  geom_point(stat = "identity", 
           position = "stack", 
           fill = "black") +
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth", 
              se = FALSE,
              color = "red") +
  theme_bw() +
  xlab("Decade") +
  ylab("Number of Tornadoes") +
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16), 
        legend.text = element_text(size = 16)) +
  scale_y_continuous(expand = c(0, 0),  
                     limits=c(0, 1850), 
                     breaks = seq(0, 1850, 250)) +
  scale_x_continuous(expand = c(0, 0), 
                     limits=c(1950, 2023), 
                     breaks = seq(1950, 2023, 10))
```




## Making figures with tmap:

```{r}
#install.packages("tmap")
library(tmap)
```

