---
`title: "Heidi_ExploringTrends"
author: "Heidi Erfourth and Zoe Schroder"
date: "2023-07-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

“Exploring trends in tornado outbreak environments and greenhouse gas concentrations” 

`Pursue publication:` International Journal of Climatology
                      https://mc.manuscriptcentral.com/joc 

**Make sure to follow the Author Resources for publication in the Journal. This includes page formatting, section headers, figure format, table format, figure/table location in manuscript, etc.**
```{r}
install.packages("RColorBrewer")
```

```{r}
suppressMessages(library("dplyr"))
suppressMessages(library("ggplot2"))
suppressMessages(library("tmap"))
suppressMessages(library("USAboundaries"))
suppressMessages(library("sf"))
suppressMessages(library("gridExtra"))
```

DATA & METHODS
```{r}
load("TornadoOutbreaks.RData")
AnnGHG <- read.csv("AnnualAverages_GHG.csv")
```

```{r}
cr <- RColorBrewer::brewer.pal(9, "Reds")
cr <- cr[-c(1:3)]
```
##Tornado Data
**  Needs (numbers for table): 
** 1. Number of Days
** 2. Number of Groups
** 3. Number of Tornadoes
** 4. Number of Casualties

##Create a figure (Map) with multiple clusters on a single day
Finding a Single Day with Multiple Clusters 
```{r}
MultiClus<-BigDays.sfdfT %>% # Cluster Data
  group_by(cDate)%>% # Group by the c. days to see if there are duplicates
  summarize(num_of_clusters = n())%>%  # Count the number of clusters on each cDate.
  arrange(desc(num_of_clusters)) # Arrange the number of clusters column in descending order.
```

Data from June 4, 1999 Outbreak
```{r}
#Group Number: 1686, 1688
#Cluster 1 Data
June4<-BigDays.sfdfT%>%
  dplyr::filter(ID=="199906041686")

June4cent <-BigDayCentroids.df %>%
  dplyr::filter(ID=="199906041686")

June4Torns <- BigDayTornadoes%>%
  dplyr::filter(ID=="199906041686")

#Cluster 2 Data
June4_1<-BigDays.sfdfT%>%
  dplyr::filter(ID=="199906041688")

June4_1cent <-BigDayCentroids.df %>%
  dplyr::filter(ID=="199906041688")

June4_1Torns <- BigDayTornadoes%>%
  dplyr::filter(ID=="199906041688")

#Combining Cluster Data
June <- rbind(June4, June4_1)#Overall Data 
Junecent <- rbind(June4cent, June4_1cent)#Centroids
JuneTorns <- rbind(June4Torns, June4_1Torns)#Tornadoes
```
 
State Boarders for Contiguous United States 
```{r}
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")] #Removes AK and HI from list of states
stateBorders <- us_states(states = sts) 
```

**Plotting June 4, 1999 Outbreak Data (Convective Day: 6 am 27th to 6 am 28th)**
```{r}
#Creating Map Projection
tm_shape(stateBorders,
          projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84
          +datum=WGS84 +units=m+no_defs")+
  
  #Formatting Borders
  tm_borders(col ="gray", 
             alpha = 1)+
  
  #Compass and Scale Bar Formatting
  tm_compass(text.size = 0.7, 
             lwd = 2, 
             color.dark = "gray")+
  tm_scale_bar(width = 0.27, 
               text.size = 0.6, 
               lwd = 1.75,  
               color.dark = "gray")+
  
  #Formatting Plot 
  tm_layout(legend.position = c(0.01,0),
            attr.position=c("right","bottom"), 
            inner.margins = c(0.3,0.2,0.3,0.2), 
            title="June 4, 1999", 
            title.position = c(0.40, 0.95))+
  
#Creating Polygons for Clusters
tm_shape(June, 
         is.master = TRUE, 
         projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 
         +units=m +no_defs" )+
  
  #Formatting Polygons
  tm_polygons(alpha = 0, 
              border.col = "black", 
              lwd = 1.5)+

#Plotting Individual Tornadoes
tm_shape(JuneTorns, 
         projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84
         +units=m +no_defs" )+
  
  #Formatting Tornado Bubbles
  tm_bubbles(size = 0.2,
             border.lwd=NA, 
             palette = cr, 
             col="mag",
             breaks=seq(0,6,by=1),
             labels=c("F0", "F1","F2", "F3", "F4", "F5"),
             title.col="Magntiude")+
  
#Plotting Centroids
tm_shape(Junecent,
         is.master=TRUE,
         projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84
         +units=m +no_defs" )+
  
  #Formatting Centroids
  tm_symbols(size = 1.25, 
             col="black", 
             shape=24)
```

RESULTS

Plotting Outbreaks for 1994-2022
```{r}
#Creating Map Projection
tm_shape(stateBorders,
         projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84
         +units=m +no_defs")+
  
  #Formatting Plot
  tm_borders(col ="gray", 
             alpha = 1)+
  tm_layout(title="Tornado Outbreaks from 1994-2022",
            title.position = c(0.3, 0.95),
            inner.margins = c(0.1,0.1,0.1,0.1))+
  
  #Formatting Compass & Scale Bar
  tm_compass(text.size = 0.7, 
             lwd = 2, 
             color.dark = "gray",
             position = c(0,0.2))+
  tm_scale_bar(width = 0.27, 
               text.size = 0.6,
               lwd = 1.75, 
               color.dark = "gray", 
               position = c(0.01,0))+
  
#Plotting Individual Clusters Using Centroids
tm_shape(BigDayCentroids.df, 
         projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 
         +units=m +no_defs" )+
  
  #Formatting Centroids
  tm_bubbles(size = 0.1,
             border.lwd=NA, 
             col = "orange3")
``` 

##Annual Change in Tornado Outbreaks

Grouping Total Number of Outbreaks per Year
```{r}
Totout_yr <- BigDays.sfdfT%>% 
  group_by(Year)%>%
  summarize(Totout_Year = n())
```

Graphing Total Outbreaks per Year
```{r}
#Plotting Outbreaks
ggplot(Totout_yr,
       aes(x=Year, y=Totout_Year))+
  
  #Line Graph
  geom_line(stat="identity", 
            color="orange3", 
            linewidth=1)+
  
  #Creating Linear Regression Line
  stat_smooth(method = "lm", 
              formula = y~x, 
              geom="smooth", 
              color="black", 
              linewidth=1)+
  
  #Plot Formatting
  ggtitle("Total Tornado Outbreaks Per Year")+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5), 
        plot.margin=margin(0.5,0.5,0.5,0.5, "cm"), 
        axis.text.x=element_text(angle=45, hjust=1))+
  
  #Axes Formatting
  xlab("Year")+
  ylab("Number of Tornado Outbreaks per Year")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022),
                     breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0,0),
                     limits=c(20,50), 
                     breaks=seq(20,50,5))
```


##Annual Averages in GHG Emissions using BigDays_1year data
CO2 Annual Average
```{r}
#Plotting CO2 1 Year Averages 
CO2_avg <- ggplot(BigDays.sfdfT, 
                  aes(x= Year, y=CO2_1year))+
  
   # #Creating a Linear Regression Line
   # stat_smooth(method = "lm", 
   #            formula = y~x, 
   #            geom="smooth", 
   #            color="black", 
   #            linewidth=1, se=F)+
  
  #Line Graph
   geom_line(stat="identity", 
             color = "dodgerblue3",
             linewidth = 1)+
  
  #Formatting Plot
   theme_bw()+
   theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting Axes
   ylab(expression("CO"[2]* "(ppm)"))+
   scale_x_continuous(expand = c(0, 0), 
                     limits=c(1994, 2022), 
                     breaks = seq(1994, 2022,4)) + 
   scale_y_continuous(expand = c(0,0), 
                     limits=c(345, 415), 
                     breaks = seq(350, 415, 10)) 
```

CH4 Annual Average
```{r}
#Plotting CH4 Data
CH4_avg <- ggplot(BigDays.sfdfT, 
                  aes(x=Year, y = CH4_1year))+
  
  # #Creating a Linear Regression Line
  #  stat_smooth(method = "lm", 
  #             formula = y~x, 
  #             geom="smooth", 
  #             color="black", 
  #             linewidth=1, se=F)+
  
  #Line Graph
   geom_line(stat="identity",
             color = "purple3",
             linewidth = 1)+
  
  #Formatting Plot
   theme_bw()+
   theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting Axes
   ylab(expression("CH"[4]* "(ppb)"))+
   scale_x_continuous(expand=c(0,0), 
                      limits=c(1994, 2022), 
                      breaks=seq(1994,2022,4))+
   scale_y_continuous(expand=c(0,0), 
                      limits=c(1720,1880), 
                      breaks=seq(1720,1880, 20))
```

O3 Annual Average
```{r}
#Plotting O3 Data
O3_avg<-ggplot(BigDays.sfdfT,
               aes(x=Year, y=O3_1year))+
  
  # #Creating a Linear Regression Line
  #  stat_smooth(method = "lm", 
  #             formula = y~x, 
  #             geom="smooth", 
  #             color="black", 
  #             linewidth=1, se=F)+
  
  #Line Graph
   geom_line(stat="identity",
             color = "goldenrod3",
             linewidth = 1)+
  
  #Formatting Plot
   theme_bw()+
   theme(axis.text.x=element_text(angle=45, hjust = 1))+
  
  #Formatting axes
   ylab(expression("O"[3]* "(DU)"))+
   scale_x_continuous(expand=c(0,0), 
                      limits=c(1994, 2022), 
                      breaks=seq(1994,2022,4))+
   scale_y_continuous(expand=c(0,0), 
                      limits=c(90,175), 
                      breaks=seq(90,175, 20))
```

AGGI Annual Average
```{r}
#Plotting AGGI Data
AGGI_avg<- ggplot(BigDays.sfdfT,
                  aes(x=Year, y=AGGI_1year))+
  
  #Creating a Linear Regression Line
   # stat_smooth(method = "lm", 
   #            formula = y~x, 
   #            geom="smooth", 
   #            color="black", 
   #            linewidth=1, se=F)+
   # 
  #Line Graph
   geom_line(stat="identity", 
             color = "green4",
             linewidth = 1)+
  
  #Formatting plot
   theme_bw()+
   theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting Axes
   ylab("AGGI")+
   xlab("Year")+
   scale_x_continuous(expand=c(0,0), 
                      limits=c(1994, 2022), 
                      breaks=seq(1994,2022,4))+
   scale_y_continuous(expand=c(0,0), 
                      limits=c(0.5,2),
                      breaks=seq(0.5,2,0.2))
```

4x4 Grid Annual Averages
```{r}
grid.arrange(CO2_avg, CH4_avg, O3_avg, AGGI_avg, nrow=2,ncol=2, top ="Annual Averages of Greenhouse Gas Emissions")
```
  
##10-Year Averages in GHG Emissions using BigDays_1year data

CO2 10-year Averages
```{r}
#Plotting CO2 Data
CO2_10 <-ggplot(BigDays.sfdfT, 
                aes(x=Year, y=CO2_10year)) +
  
  # #Creating a Linear Regression Line
  #  stat_smooth(method = "lm", 
  #             formula = y~x, 
  #             geom="smooth", 
  #             color="black", 
  #             linewidth=1,
  #             se=F)+
  
  #Line Graph
  geom_line(stat="identity", 
            color = "dodgerblue3",
            linewidth = 1)+
  
  #Formatting Plot
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting axes
  xlab("Year") +
  ylab(expression("CO"[2]*"(ppm)"))+
  scale_x_continuous(expand = c(0, 0), 
                     limits=c(1994, 2022), 
                     breaks = seq(1994, 2022, 10)) + 
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(345, 415), 
                     breaks = seq(350, 415, 10)) 
```

CH4 10 year Averages
```{r}
#Plotting CH4 Data
CH4_10<-ggplot(BigDays.sfdfT, 
               aes(x=Year, y=CH4_10year)) +
  
  # #Creating a Linear Regression Line
  #  stat_smooth(method = "lm", 
  #             formula = y~x, 
  #             geom="smooth", 
  #             color="black", 
  #             linewidth=1,
  #             se=F)+
  
  #Line Graph
  geom_line(stat="identity",
            color="purple3",
            linewidth=1 )+
  
  #Formatting Plot
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting Axes
  xlab("Year")+
  ylab(expression("CH"[4]* "(ppb)"))+
  scale_x_continuous(expand=c(0,0), 
                     limits=c(1994, 2022), 
                     breaks=seq(1994,2022,10))+
  scale_y_continuous(expand=c(0,0), 
                     limits=c(1720,1880), 
                     breaks=seq(1720,1880, 20))
```

O3 10-year Averages
```{r}
#Plotting O3 Data
O3_10<-ggplot(BigDays.sfdfT, 
              aes(x=Year, y=O3_10year)) +
  
  # #Creating a Linear Regression Line
  #  stat_smooth(method = "lm", 
  #             formula = y~x, 
  #             geom="smooth", 
  #             color="black", 
  #             linewidth=1,
  #             se=F)+
  
  #Line Graph
  geom_line(stat="identity", 
            color="goldenrod3",
            linewidth=1 )+
  
  #Formatting Plot
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting Axes
  xlab("Year")+
  ylab(expression("O"[3]*"(DU)"))+
  scale_x_continuous(expand=c(0,0), 
                     limits=c(1994, 2022),
                     breaks=seq(1994,2022,10))+
  scale_y_continuous(expand=c(0,0), 
                     limits=c(90,150), 
                     breaks=seq(90,150, 10))
```

AGGI 10-year Averages
```{r}
#Plotting AGGI Data
AGGI_10<-ggplot(BigDays.sfdfT, 
                aes(x=Year, y=AGGI_10year)) +
  
  # #Creating a Linear Regression Line
  #  stat_smooth(method = "lm", 
  #             formula = y~x, 
  #             geom="smooth", 
  #             color="black", 
  #             linewidth=1,
  #             se=F)+
  
  #Line Graph
  geom_line(stat="identity", 
            color="green4",
            linewidth=1 )+
  
  #Formatting Plot
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  
  #Formatting Axes
  xlab("Year")+
  ylab("AGGI")+
  scale_x_continuous(expand=c(0,0), 
                     limits=c(1994, 2022), 
                     breaks=seq(1994,2022,10))+
  scale_y_continuous(expand=c(0,0), 
                     limits=c(0.5,2),
                     breaks=seq(0.5,2,0.2))
```

```{r}
grid.arrange(CO2_10, CH4_10, O3_10, AGGI_10, nrow=2, top ="10-Year Averages of Greenhouse Gas Emissions")
```


**Seasonal Change in Tornado Outbreaks**

Spatial Density by Month ((I need help with the filtering))
Needs:
-Filter data for each month
-Calculate the total number of outbreaks per month
-create a figure for each month similar to the total outbreaks map
-Comment on which ones are the most dense and where
```{r}
BigDayCentroids.df%>%
  group_by(Month)%>%
  summarise(out= n())
```

**Relationship between GHG Emission and Total Number of Outbreaks**
1 year Correlation Matrix
```{r}
#Gathering Data
test_1yr <- BigDays.sfdfT %>%
  
  #Filtering Data
  group_by(Year) %>%
  summarize(Totout_yr =n(), #Total Number of outbreaks
            avgTorn = mean(nT), #Average number of outbreak tornadoes per year
            totTorn = sum(nT), #Total number of outbreak tornadoes per year
            totCas = sum(GroupDayCas), #Total number of casualties by year
           
            #Total Number of Tornadoes Per Rating
            T_F0 = sum(n0), 
            T_F1=sum(n1),  
            T_F2=sum(n2), 
            T_F3=sum(n3), 
            T_F4=sum(n4), 
            T_F5=sum(n5),
            
            #Maximum Emissions for 1 Year
            CO2_1yr = max(CO2_1year),
            CH4_1yr = max(CH4_1year),
            O3_1yr = max(O3_1year),
            AGGI_1yr = max(AGGI_1year)) %>%
  
  #Removing Geometry Column
  st_drop_geometry()

#Correlation Matrix
cor(test_1yr)

test_1yr
```

10 year Correlation Matrix
```{r}
#Gathering Data
test_10yr <- BigDays.sfdfT %>%
  #Grouping Data
  group_by(Year) %>%
  summarize(Totout_yr =n(), #Total Number of outbreaks
            avgTorn = mean(nT), #Average number of outbreak tornadoes per year
            totTorn = sum(nT), #Total number of outbreak tornadoes per year
            totCas = sum(GroupDayCas), #Total number of casualties by year
            
            #Total number of Tornadoes Per Rating
            T_F0 = sum(n0), 
            T_F1=sum(n1),  
            T_F2=sum(n2), 
            T_F3=sum(n3), 
            T_F4=sum(n4), 
            T_F5=sum(n5),
            
            #Maximum Emissions for 10 years
            CO2_10yr = max(CO2_10year),
            CH4_10yr = max(CH4_10year),
            O3_10yr = max(O3_10year),
            AGGI_10yr = max(AGGI_10year)) %>%
  
  #Removing Geometry Column
  st_drop_geometry()

#Correlation Matrix
cor(test_10yr) 
```

**Line Graph for 1-year & 10-year emissions with Tornado Outbreak Data**
```{r}
#Gathering Data
GHG_Out <-BigDays.sfdfT%>%
  dplyr::select(Year, CO2_1year, CH4_1year, O3_1year, AGGI_1year, nT, GroupDayCas)%>%
  #Grouping Data
  group_by(Year)%>%
  summarise(Totout_yr=n(),
            T_avgyear = mean(nT),
            T_Cas = sum(GroupDayCas))
```

#1 Year Averages#
1 Year Averages CO2 and Outbreak Relationship
```{r}
ggplot(GHG_Out)+
  
  #Plotting CO2 1 year Average
  geom_line(aes(x=Year, 
                y=CO2_1year), 
            stat= "identity",
            color="dodgerblue3", 
            linewidth=1)+
  
  #Plotting Total # of Outbreak per year
  geom_bar(aes(x=Year, y=Totout_yr*8), 
            stat="identity", 
            fill="orange3", 
            linewidth=1)+
  
  #Formatting plot
  theme_bw()+
  
  #Formatting Axes
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022), 
                     breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(0, 415),
                     breaks = seq(0,450, 50), 
                     sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks"))
```

1 Year Averages CH4 and Outbreak 
```{r}
CH4_1<-ggplot(GHG_Out)+
  
  #Plotting CH4 1 year averages
  geom_line(aes(x=Year, y=(Avg1_CH4)), 
            color="purple3", 
            linewidth=1)+
  
  #Plotting Total # of Outbreaks  per year
  geom_bar(aes(x=Year, y=T_Year*25), 
           fill="orange3", 
           linewidth=1)+
  
  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("CH4 Emissions (ppb)")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022),
                     breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(0,3000), 
                     breaks = seq(0,3000, 500), 
                     sec.axis = sec_axis(~./25, name = "Total Tornado Outbreaks"))
```

1 Year Averages O3 and Outbreak Relationship
```{r}
O3_1<-ggplot(GHG_Out)+
  
  #Plotting O3 1 year averages
  geom_line(aes(x=Year, y=(Avg1_O3)), 
            color="goldenrod3", 
            linewidth=1)+
  
  #Plotting Total # of outbreaks per year
  geom_bar(aes(x=Year, y=T_Year/.5), 
           fill="orange3", 
           linewidth=1)+
  
  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("O3 Emissions (DU)")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022),
                     breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0),
                     limits=c(0,200), 
                     breaks = seq(0,200,20), 
                     sec.axis = sec_axis(~.*.5, name = "Total Tornado Outbreaks"))
```

1 Year Averages AGGI and Outbreak Relationship
```{r}
AGGI_1<-ggplot(GHG_Out)+
  
  #Plotting AGGI 1 year averages
  geom_line(aes(x=Year, y=(Avg1_AGGI)),
            color="green4", 
            linewidth=1)+
  
  #Plotting Total # of outbreaks per year
  geom_bar(aes(x=Year, y=T_Year/25), 
           color="orange3", 
           linewidth=1)+
  
  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("AGGI")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022), 
                     breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0,0), 
                     limits=c(0,2),
                     breaks = seq(0,2,.1), 
                     sec.axis = sec_axis(~.*25, name = "Total Tornado Outbreaks"))
```

```{r}
grid.arrange(CO2_1, CO2_1, CO2_1, CO2_1, nrow=2, top ="Annual Average Emissions & Total Outbreaks per Year")
```

#10 Year Averages
10 year averages CO2 and Tornado Outbreaks
```{r}
CO2_10<-ggplot(GHG_Out)+
  
  #Plotting 10 Year average CO2
  geom_line(aes(x=Year, y=(Avg10_CO2)), 
            stat= "identity",
            color="dodgerblue3", 
            linewidth=1)+
  
  #Plotting Total # of outbreaks per year
  geom_bar(aes(x=Year, y=T_Year*8), 
           stat="identity", 
           fill="orange3", 
           linewidth=1)+
  
  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022), 
                     breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(0, 415), 
                     breaks = seq(0,415, 50), 
                     sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks")) 
```

10 year averages CH4 and Tornado Outbreaks
```{r}
CH4_10<-ggplot(GHG_Out)+
  
  #Plotting CH4 10 year averages
  geom_line(aes(x=Year, y=(Avg10_CH4)), 
            color="purple3", 
            linewidth=1)+
  
  #Plotting Total # of outbreaks per year
  geom_bar(aes(x=Year, y=T_Year*25), 
           fill="orange3", 
           linewidth=1)+

  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("CH4 Emissions (ppb)")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022), 
                     breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(0, 3000),
                     breaks = seq(0,3000, 500), 
                     sec.axis = sec_axis(~./25, name = "Total Tornado Outbreaks")) 
```

10 year averages O3 and Tornado Outbreaks
```{r}
O3_10<-ggplot(GHG_Out)+
  
  #Plotting 10 year averages of AGGI
  geom_line(aes(x=Year, y=(Avg10_O3)), 
            color="goldenrod3", 
            linewidth=1)+
  
  #Plotting Total # of outbreaks per year
  geom_bar(aes(x=Year, y=T_Year/.5), 
           fill="orange3", 
           linewidth=1)+
  
  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("O3 Emission (DU)")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022), 
                     breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(0, 200), 
                     breaks = seq(0,200, 20), 
                     sec.axis = sec_axis(~.*.5, name = "Total Tornado Outbreaks")) 
```

10 year averages AGGI and Tornado Outbreaks
```{r}
AGGI_10<-ggplot(GHG_Out)+
  
  #Plotting 10 year averages AGGI
  geom_line(aes(x=Year, y=(Avg10_AGGI)), 
            color="green4", 
            linewidth=1)+
  
  #Plotting Total # of outbreaks per year
  geom_bar(aes(x=Year, y=T_Year/25), 
           fill="orange3", 
           linewidth=1)+
  
  #Formatting Plot
  theme_bw()+
  
  #Formatting Axes
  ylab("AGGI")+
  scale_x_continuous(expand = c(0,0),
                     limits=c(1994,2022), 
                     breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(0, 2), 
                     breaks = seq(0,2,.5), 
                     sec.axis = sec_axis(~.*25, name = "Total Tornado Outbreaks")) 
```

**need to make a 4x4 Grid**




##C02 Trend Analysis with Total Casualties
1 year average of CO2 & Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg1_CO2)), stat= "identity",color="dodgerblue3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/20), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~.*20, name = "Total Casualties"))
```

10 year averages CO2 and Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg10_CO2)), stat= "identity",color="dodgerblue3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/20), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~.*20, name = "Total Casualties")) 
```


##CH4 Trend Analysis with Total Casualties
1 Year Averages CH4 and Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg1_CH4)), stat= "identity",color="firebrick3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/2), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("CH4 Emissions (ppb)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), limits=c(0,3000), breaks = seq(0,3000, 500), sec.axis = sec_axis(~.*2, name = "Total Casualties"))
  

```

10 year averages CH4 and Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg10_CH4)), stat= "identity",color="firebrick3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/2), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("CH4 Emissions (ppb)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 3000), breaks = seq(0,3000, 500), sec.axis = sec_axis(~.*2, name = "Total Casualties")) 
```


##O3 Trend Analysis with Total Casualties
1 Year Averages O3 and Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg1_O3)), stat= "identity",color="darkorange3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/30), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("O3 Emissions (DU)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), limits=c(0,200), breaks = seq(0,200,20), sec.axis = sec_axis(~.*30, name = "Total Casualties"))
```

10 year averages O3 and Tornado Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg10_O3)), stat= "identity",color="darkorange3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/30), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("O3 Emission (DU)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 200), breaks = seq(0,200, 20), sec.axis = sec_axis(~.*30, name = "Total Casualties")) 
```


##AGGI Trend Analysis with Total Casualties
1 Year Averages AGGI and Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg1_AGGI)), stat= "identity",color="green4", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/4000), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("AGGI")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0,0), limits=c(0,2), breaks = seq(0,2,.1), sec.axis = sec_axis(~.*4000, name = "Total Casualties"))
```

10 year averages AGGI and Total Casualties
```{r}
ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg10_AGGI)), stat= "identity",color="green4", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas/4000), stat="identity", fill="deeppink3", linewidth=1)+
  theme_bw()+
  ylab("AGGI")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 2), breaks = seq(0,2,.5), sec.axis = sec_axis(~.*4000, name = "Total Casualties"))
```


