---
title: "Heidi_ExploringTrends"
author: "Heidi Erfourth and Zoe Schroder"
date: "2023-07-25"
output: html_document
---

“Exploring trends in tornado outbreak environments and greenhouse gas concentrations” 

`Pursue publication:` International Journal of Climatology
                      https://mc.manuscriptcentral.com/joc 

**Make sure to follow the Author Resources for publication in the Journal. This includes page formatting, section headers, figure format, table format, figure/table location in manuscript, etc.**

```{r}
# Do not need to do these everytime. Once installed on your computer you are finished. 
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("tidyverse")
#install.packages("tmap")
#install.packages("devtools")
#devtools::install_github("ropensci/USAboundaries")
#devtools::install_github("ropensci/USAboundariesData")
#install.packages("gridExtra")
```

```{r}
library("dplyr")
library("ggplot2")
library("tmap")
library("USAboundaries")
library("sf")
library("gridExtra")
```

DATA & METHODS
Tornado Clusters 
  Setup Data to produce a clustering with Space-Time Distances

  Needs (numbers for table): 
  1. Number of Days
  2. Number of Groups
  3. Number of Tornadoes
  4. Number of Casualties

`You need to load the data before you can do anything. Do this:`
```{r}
load("TornadoOutbreaks.RData")
```


Cluster: >= 10 tornadoes `This is already done for you. This is the BigDays.sfdfT data`
```{r}
BigDays.sfdfT[,-80]%>% #What is the -80 for? I kept getting an error from the geometry column -80 tells it not to read that column
  group_by(cDate)%>%
  filter(nT>=10)
```
  
  Create a figure (Map) with multiple clusters on a single day
  
`If you are wanting to find a day that has more than one cluster, do this: `
```{r}
BigDays.sfdfT %>% # Cluster Data
  group_by(cDate) %>% # Group by the convective days to see if there are duplicates.
  summarize(num_of_clusters = n()) %>% # Count the number of clusters on each cDate.
  arrange(desc(num_of_clusters)) # Arrange the number of clusters column in descending order. 
```
  
  
  Things that need to be accomplished with said figure
  - creating clustering `This is already done for you. You are doing the same thing that I did in my code. Follow that as an example. You just need to pick which outbreak you want to show.`
  - creating a line around the clustering area `The clustering area is already computed. If you plot the specific outbreak you want from the BigDays.sfdfT it will plot the clustering area. Again reference my code for a figure similar to this.`

## Greenhouse Gas Emissions Data
  Value Selection for each Lag (Table Values)
  Needs:
  Average Emissions per year for CO2, CH4, SF6, N2O, and O3

```{r}
#need to read in the data for the excel file
AnnGHG <- read.csv("AnnualAverages_GHG.csv")

#validating that data looks correct and is the right dimensions
# dim(AnnGHG)
# str(AnnGHG)
```

*Here is the code to rename a column in R: * Fixing the typo in the excel file zero to O
```{r}
colnames(AnnGHG)[colnames(AnnGHG) == "N20_ppb"] = "N2O_ppb"
```

Calculating the Average Emissions per Year (for Table)
```{r}
AnnGHG %>%
  group_by(year) %>%
  summarize(avgCO2 = mean(CO2_ppm, na.rm = TRUE),
            avgCH4 = mean(CH4_ppb, na.rm = TRUE), 
            avgSF6 = mean(SF6_ppt, na.rm=TRUE), 
            avgN2O =mean(N2O_ppb, na.rm=TRUE), 
            avgO3 =mean(O3_DU, na.rm=TRUE))
```

## RESULTS
Spatial Distribution of Tornado Outbreaks in Contiguous US 
  
Needs:
- Map of US with compass and scale
- Plotting Outbreaks (NOT individual Tornadoes)

NOTE: These are only for the years 1994 to 2022. Moving forward, you will only use CO2, CH4, O3, and AGGI. N2O and SF6 are excluded because the data does not extend back to 1994. 

```{r}
#getting State Borders for contiguous United States
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")] #Removes AK and HI from list of states
stateBorders <- us_states(states = sts) 
```

```{r}
#creating map with outbreak data
tm_shape(stateBorders,is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m   +no_defs")+
  tm_borders(col ="gray", alpha = 1)+
  tm_compass(text.size = 1, 
             lwd = 2, 
             color.dark = "gray")+
  tm_scale_bar(width = .3, 
               text.size = 0.8, 
               lwd = 1.75, 
               color.dark = "gray")+
  tm_layout(title="Tornado Outbreaks from 1994-2022", title.position=c("center", "top"), attr.position = c("left","bottom"), inner.margins = c(.1, .1, .1, .1))+
tm_shape(BigDayCentroids.df, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" )+
  tm_bubbles(size = 0.1, 
             col = "orchid4")


``` 

Annual Change in Tornado Outbreaks
Needs:
- Year
- Total count of Outbreak within that year

```{r}
#creating grouping for number of outbreaks per year
T_yr <- BigDays.sfdfT%>% 
  group_by(Year)%>%
  summarize(T_Year = n())

#creating a line graph for T-outbreaks per year (probably won't need this one, but I did both incase we wanted to pick one or the other)
ggplot(T_yr)+
  geom_line(aes(x=Year, y=T_Year), color="orchid4", linewidth=1)+
  ggtitle("Total Tornado Outbreaks Per Year")+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Year")+
  ylab("Number of Tornado Outbreaks per Year")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0,0),limits=c(20,50),  breaks=seq(20,50,5))
```

```{r}
#creating a bar graph with regression line for T-Outbreaks per year
ggplot(T_yr, aes(x=Year, y=T_Year))+
  geom_bar(stat="identity",position = "stack",  fill="orchid4", width=0.9)+
  stat_smooth(method = "lm", formula = y~x, geom="smooth", color="black", linewidth=1.5)+
  ggtitle("Total Outbreaks Per Year")+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Year")+
  ylab("Number of Tornado Outbreaks")+
  scale_x_continuous(expand=c(0,0), limits=c(1994,2022), breaks=c(seq(1994, 2022, 2)))+
  scale_y_continuous(expand=c(0,0), limits=c(0,50), breaks=c(seq(0,50,5)))

```
*slightly positive regression line, therefore outbreaks are on the rise.*

Annual Change in GHG Emissions
Needs:
-Year
-Emission Values for that Year
-Regression Line showing overall trend

*creating individual annual trend charts for CO2, CH4, O3, AGGI*

CO2 Annual Trend Analysis Chart
```{r}  
AnnCO2 <- ggplot(AnnGHG, aes(x=year, y=CO2_ppm))+
  geom_line(stat="identity", color="dodgerblue3", linewidth=1)+
  #ggtitle(bquote("Annual Average of Carbon Dioxide CO"[2]))+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Years Recorded")+
  ylab("CO2 (ppm)")+
  scale_x_continuous(expand=c(0,0), limits=c(1955,2025), breaks=c(seq(1960,2025,10)))+
  scale_y_continuous(expand=c(0,0), limits=c(300,450), breaks=c(seq(300,450,30)))
```
  *Positive regression line, values are on the rise*
  
CH4 Annual Trend Analysis Chart
```{r}  
AnnCH4<-ggplot(AnnGHG, aes(x=year, y=CH4_ppb))+
  geom_line(stat="identity",color="firebrick3", linewidth=1)+
 # ggtitle(bquote("Annual Average of Methane CH"[4]))+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Years Recorded")+
  ylab(bquote("CH4 (ppb)"))+
  scale_x_continuous(expand=c(0,0), limits=c(1980,2025), breaks=c(seq(1980, 2025, 10)))+
  scale_y_continuous(expand=c(0,0), limits=c(1600,2000), breaks=c(seq(1600,2000,100)))
```
  *Positive regression line, values are on the rise*
  
O3 Annual Trend Analysis Chart  
```{r}
#DU = Dobson Units
AnnO3<-ggplot(AnnGHG, aes(x=year, y=O3_DU))+
  geom_line(stat="identity",color="darkorange2", linewidth=1)+
  #stat_smooth(method = "lm", formula = y~x, geom="smooth", color="black", linewidth=1.5)+
  #ggtitle(bquote("Annual Average of Ozone O"[3]))+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Years Recorded")+
  ylab("03 (DU)")+
  scale_x_continuous(expand=c(0,0), limits=c(1975,2025), breaks=c(seq(1975, 2020, 10)))+
  scale_y_continuous(expand=c(0,0), limits=c(50, 250), breaks=c(seq(50,250, 50)))
```
  *Negative regression line, values are decrease. However, has the most variation throughout the years*
  
AGGI Annual Trend Analysis Chart
```{r}
AnnAGGI<-ggplot(AnnGHG, aes(x=year, y=AGGI))+
  geom_line(stat="identity",color="green4", linewidth=1)+
  #ggtitle("Annual Average for Annual Greenhouse Gas Index (AGGI)")+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Years Recorded")+
  ylab("AGGI")+
  scale_x_continuous(expand=c(0,0), limits=c(1975,2025), breaks=c(seq(1975, 2020, 10)))+
  scale_y_continuous(expand=c(0,0), limits=c(0.5,2), breaks=c(seq(0.5,2,0.3)))

```
 *Positive regression line, values are on the rise*

combining annual avearges into a 4x4 grid
```{r}
grid.arrange(AnnCO2, AnnCH4, AnnO3, AnnAGGI, nrow = 2, top=("Emissions Annual Averages"))
```
 `This plot is fine for showing the trends with regards to all data that is available. No value when comparing to the Tornado Outbreaks.`

*Combining the Annual Average Charts into a 4x4 grid*

`make these plots as line graphs using ggplot like you did above. Keep the colors consistent for the GHG variables. Make regression line black.`
```{r}
CO2_avg <- qplot(Year, CO2_1year, data = BigDays.sfdfT)+
  stat_smooth(method = "lm", formula = y~x, geom = "smooth", color="darkturquoise", linewidth=1.5)+
  theme_bw()+
  ylab("Average of CO2")
CH4_avg <- qplot(Year, CH4_1year, data = BigDays.sfdfT)+
  stat_smooth(method = "lm", formula = y~x, geom = "smooth", color="plum3", linewidth=1.5)+
  theme_bw()+
  ylab("Average of CH4")
O3_avg <-  qplot(Year, O3_1year, data = BigDays.sfdfT)+
  stat_smooth(method = "lm", formula = y~x, geom = "smooth", color="pink3", linewidth=1.5)+
  theme_bw()+
  ylab("Average of O3")
AGGI_avg<- qplot(Year, AGGI_1year, data= BigDays.sfdfT)+
  stat_smooth(method = "lm", formula = y~x, geom = "smooth", color="seagreen3", linewidth=1.5)+
  theme_bw()+
  ylab("Average of AGGI")

grid.arrange(CO2_avg, CH4_avg, O3_avg, AGGI_avg, nrow=2,ncol=2, top ="Annual Averages of Greenhouse Gas Emissions")
```
  *Slightly confused here because the regression line of O3 is positive when the regression line for the individual one was negative. The 4x4 is using the BigDays Data because I tried using the AnnGHG but it kept giving me a mapping error*
  
`The figures that you did previously is for all data prior to 1994. The figures that you did in this code chunk are for 1993 to 2021 (1-year prior avg so 1993 data for 1994 avg, 1994 data for 1995 avg, etc.). This is why there is a difference. For 1-year average values, you only want to look at 1993 to 2022. For 5-year average, you want 1989 to 2021. For 10-year average, you want 1983 to 2021. So, if you remove all years prior to 1993 above, the regression line flips to a positive trend. For this research, you should focus only on the values in the BigDays.sfdfT when comparing it to the tornado data.`

10-year averages (Table)
*The 10 year averages are already calculated for you. They are in the BigDays.sfdfT file found in the TornadoOutbreaks.RData*
**The avg for 1994 is 1984 - 1994 not inclusive. Meaning it includes the data for years January 1984 through December 1993.**

Need to get the the 10 year values for the 90s, 00s, 10s, and 20s for the table
*I don't think I need the charts below*


CO2 10-year averages trend
```{r}
CO2_10 <-ggplot(BigDays.sfdfT) + 
  geom_line(aes(x = Year, y = CO2_10year), 
            color = "dodgerblue3",
            linewidth = 1)  +
  theme_bw() +
  xlab("Year") +
  ylab("10-year Average CO2 Emissions") +
  theme(axis.text = element_text(size = 12), 
        axis.text.x = element_text(angle = 45, 
                                   vjust = 1.2, 
                                   hjust = 1.2),
        axis.title = element_text(size = 14), 
        legend.text = element_text(size = 14),
        ) +
  scale_x_continuous(expand = c(0, 0), 
                     limits=c(1983, 2021), 
                     breaks = seq(1983, 2021, 2)) + 
  scale_y_continuous(expand = c(0, 0), 
                     limits=c(345, 415), 
                     breaks = seq(350, 415, 10)) 
```

CH4 10 year averages trend
```{r}
CH4_10<-ggplot(BigDays.sfdfT) +
  geom_line(aes(x=Year, y= CH4_10year), color="firebrick3",linewidth=1 )+
  theme_bw()+
  xlab("Year")+
  ylab("10-Year average CH4 Emissions [ppb]")+
  scale_x_continuous(expand=c(0,0), limits=c(1983, 2021), breaks=seq(1983,2021,2))+
  scale_y_continuous(expand=c(0,0), limits=c(1720,1880), breaks=seq(1720,1880, 20))

```

O3 10-year averages trend
```{r}
O3_10<-ggplot(BigDays.sfdfT) +
  geom_line(aes(x=Year, y= O3_10year), color="darkorange3",linewidth=1 )+
  theme_bw()+
  xlab("Year")+
  ylab("10-Year average O3 Emissions [DU]")+
  scale_x_continuous(expand=c(0,0), limits=c(1994, 2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand=c(0,0), limits=c(90,150), breaks=seq(90,150, 20))
  
```

AGGI 10-year averages trend
```{r}
AGGI_10<-ggplot(BigDays.sfdfT) +
  geom_line(aes(x=Year, y= AGGI_10year), color="green4",linewidth=1 )+
  ggtitle("Ten year Average of \n Annual Greenhouse Gas Index (AGGI)")+
  theme_bw()+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Year")+
  ylab("10-Year average AGGI Emissions")+
  scale_x_continuous(expand=c(0,0), limits=c(1994, 2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand=c(0,0), limits=c(0.5,2), breaks=seq(0.5,2,0.1))
  
```

`Make a 2x2 panel of the 10-year averages and a 2x2 panel of the 5-year averages. They may not be included in the paper, but you should still look at them.`

```{r}
grid.arrange(CO2_10, CH4_10, O3_10, AGGI_10, nrow=2)

```


Seasonal Change in Tornado Outbreaks

Spatial Density by Month
Needs:
-Filter data for each month
-Calculate the total number of outbreaks per month
-create a figure for each month similar to the total outbreaks map
-Comment on which ones are the most dense and where

Relationship between GHG Emission and Total Number of Outbreaks

Correlation Matrix TABLE between GHG and Total Number of Outbreaks
Needs for all four: 
-Tests for correlation
-Tests for Significance
# ```{r}
# #attempting correlation matrix
# Outbreaks <- BigDays.sfdfT%>%
#   dplyr::select(Year,CO2_1year,CH4_1year,O3_1year,AGGI_1year) 
# Outbreaks<-st_drop_geometry(Outbreaks)
# cor(Outbreaks)
 
# ```

`Try this: `
1 year emission correlation to tornado data
```{r}
test_1yr <- BigDays.sfdfT %>%
  group_by(Year) %>%
  summarize(numClus = n(), #Total number of clusters
            avgTorn = mean(nT), #Average number of outbreak tornadoes per year
            totTorn = sum(nT), #Total number of outbreak tornadoes per year
            totCas = sum(GroupDayCas), #Total number of casualties by year
            CO2_1year = max(CO2_1year),
            CH4_1year = max(CH4_1year),
            O3_1year = max(O3_1year),
            AGGI_1year = max(AGGI_1year)) %>%
  st_drop_geometry()

cor(test_1yr) 
```
`You should do this for 5-year and 10-year as well.`

5 year emission correlation to tornado data
```{r}
test_5yr <- BigDays.sfdfT %>%
  group_by(Year) %>%
  summarize(numClus = n(), #Total number of clusters
            avgTorn = mean(nT), #Average number of outbreak tornadoes per year
            totTorn = sum(nT), #Total number of outbreak tornadoes per year
            totCas = sum(GroupDayCas), #Total number of casualties by year
            CO2_5year = max(CO2_5year),
            CH4_5year = max(CH4_5year),
            O3_5year = max(O3_5year),
            AGGI_5year = max(AGGI_5year)) %>%
  st_drop_geometry()

#need to fix the tornado data for 5 year avg
cor(test_5yr) 
```

Line Graph for total outbreaks and all four variables
Needs:
-Use data for everything from 1994 to 2022
`Great Job!`
```{r}
#attempting to start this
GHG_Out <-BigDays.sfdfT%>%
  group_by(Year)%>%
  summarize(Avg1_CO2=mean(CO2_1year),
            Avg5_CO2=mean(CO2_5year),
            Avg10_CO2=mean(CO2_10year), 
            T_Year=n(),
            T_Cas = sum(GroupDayCas)) 

```

1 Year Averages CO2 and Outbreak Relationship
```{r}
#Not this one: 
#ggplot(GHG_Out)+
#  geom_bar(aes(x=Year, y=(Avg1_CO2)), stat= "identity",fill="lightblue", linewidth=1)+
#  geom_line(aes(x=Year, y=T_Year*8), stat="identity", color="blue", linewidth=1)+
  #Remove this: stat_smooth(aes(x=Year,y=T_Year*8),method = "lm", formula = y~x, geom="smooth", color="black", linewidth=1.5)+
#  theme_bw()+
#  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
#   scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks"))



ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg1_CO2)), stat= "identity",color="dodgerblue3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Year*8), stat="identity", fill="orchid4", linewidth=1)+
  theme_bw()+
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks"))
  

```
*Both have positive regressions, but T-outbreaks is a lot slower*

5 year averages CO2 and Tornado Outbreaks
```{r}
#Not this one!
#ggplot(GHG_Out)+
#  geom_bar(aes(x=Year, y=(Avg5_CO2)), stat= "identity",fill="lightblue", linewidth=1)+
#  geom_line(aes(x=Year, y=T_Year*8), stat="identity", color="blue", linewidth=1)+
#  stat_smooth(aes(x=Year,y=T_Year*8),method = "lm", formula = y~x, geom="smooth", color="black", linewidth=1.5)+
#  theme_bw()+
#  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
#   scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks"))

ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg5_CO2)), stat= "identity",color="dodgerblue3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Year*8), stat="identity", fill="orchid4", linewidth=1)+
  theme_bw()+
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks")) #Change the left y-axis label

```
 
10 year averages CO2 and Tornado Outbreaks
```{r}
#Not this one:
#ggplot(GHG_Out)+
#  geom_bar(aes(x=Year, y=(Avg10_CO2)), stat= "identity",fill="lightblue", linewidth=1)+
#  geom_line(aes(x=Year, y=T_Year*8), stat="identity", color="blue", linewidth=1)+
#  stat_smooth(aes(x=Year,y=T_Year*8),method = "lm", formula = y~x, geom="smooth", color="black", linewidth=1.5)+
#  theme_bw()+
#  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
#   scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks"))

ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg10_CO2)), stat= "identity",color="dodgerblue3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Year*8), stat="identity", fill="orchid4", linewidth=1)+
  theme_bw()+
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,5))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks")) 
```

`You can do the same for all GHG variables. You can also do all of that again for the total number of casualties (This will help with significance/importance of your paper.) and average number of tornadoes per outbreak by year. I know this is a lot of figures, but this is how you need to do trend analysis to see what the data is showing you.`

```{r}

ggplot(GHG_Out)+
  geom_line(aes(x=Year, y=(Avg1_CO2)), stat= "identity",color="dodgerblue3", linewidth=1)+
  geom_bar(aes(x=Year, y=T_Cas*8), stat="identity", fill="orchid4", linewidth=1)+
  theme_bw()+
  ylab("CO2 Emissions (ppm)")+
  scale_x_continuous(expand = c(0,0),limits=c(1994,2022), breaks=seq(1994,2022,2))+
  scale_y_continuous(expand = c(0, 0), limits=c(0, 415), breaks = seq(0,415, 50), sec.axis = sec_axis(~./8, name = "Total Tornado Outbreaks"))
  

```